---
title: "D3 tutorial for R users - barplot"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(shiny)
library(purrr)
```

<script src="www/d3.min.js" />

```{js echo=FALSE}
const initSvg = (id) => {
  const my_svg = d3
    .select(id)
    .append("svg")
    .attr("width", 500)
    .attr("height", 300);
  return my_svg;
}
const addDummyRect = (svg, scaleX = (x) => x) => {
  svg.selectAll(".my-bar-class")
    .data(dummy_data)
    .join("rect")
      .attr("x", 0)
      .attr("height", 300 / 3 * 0.8)
      .attr("y", (_, i) => 300 / 3 * i)
      .attr("width", (row) => scaleX(row.value));
}
```

In this tutorial I will prove that making a chart with D3.js is super easy. This tutorial is addressed to R users, so I will use R to demistify D3 by showing how we can achieve similar result using R.

# We need some svg.

First, we must create an `svg` element with some width and height.

We can achive it just with HTML:

```{HTML}
<div id="plot-parent">
  <svg width="500" height="300"></svg>
</div>
```

With D3.js we would select our plot parent element and append `svg` to it:

```{js eval=FALSE}
const my_svg = d3
  .select("#plot-parent")
  .append("svg")
  .attr("width", 500)
  .attr("height", 300);
```

In Shiny it is just:

```{r eval=FALSE}
div(
  id = "plot-parent",
  tags$svg(width = 500, height = 300)
)
```

Now we need to add some margin, scales, axes... Ok let's just put some bars to not get bored to much!

We will use following dummy data:

```{r echo=FALSE}
dummy_data <- data.frame(
  category = c("A", "B", "C"),
  value = c(300, 100, 50)
)
knitr::kable(dummy_data)
```

We can add bars with following code:

```{js eval=FALSE}
svg_my.selectAll(".my-bar-class")
  .data(dummy_data)
  .join("rect")
    .attr("x", 0)
    .attr("height", 300 / 3 * 0.8)
    .attr("y", (_, i) => 300 / 3 * i)
    .attr("width", (row) => row.value);
```

<div id="plot-parent"></div>

```{js echo=FALSE}
let dummy_data = [
  { category: "A", value: 300 },
  { category: "B", value: 100 },
  { category: "C", value: 50 }
];

let my_svg = initSvg("#plot-parent");
addDummyRect(my_svg);
```

*"Ok... we have 3 ugly rectangles and some spaghetti code that is not easy at all."*
Yee, looks a bit strange. But try to think about what are we doing rather then about code syntax. 
I will do the same using R:

```{r}
tags$svg(
  width = 500, height = 300,
  map(1:nrow(dummy_data), function(i) {
    tags$rect(
      x = 0,
      height = 300 / 3 * 0.8,
      y = 300 / 3 * (i - 1),
      width = dummy_data[i, ]$value
    )
  })
)
```

Ok. Now let's discuss D3.js code:

```{js eval=FALSE}
my_svg.selectAll(".my-bar-class")
```

* since there are no elements with class ".my-bar-class" in our svg, D3 creates placeholders for our data and elements

```{js eval=FALSE}
  .data(dummy_data)
```

* join `dummy_data` to placeholders

```{js eval=FALSE}
  .join("rect")
```

* append 3 `<rect>` elements (1 for each data row)

```{js eval=FALSE}
    .attr("x", 0)
```

* our bars starts on x = 0

```{js eval=FALSE}
    .attr("height", 300 / 3 * 0.8)
```

* height is calculated by dividing plot height by number of bars and multiply by 0.8 to give some padding

```{js eval=FALSE}
    .attr("y", (row, i) => 300 / 3 * i)
```

* in `attr` method we can iterate over data that we joined earlier with `data` method, using function - first argument is row data, second is row number;
  I use *arrow function* for brevity but we can use regular function - `function(row, i) { ... }`

```{js eval=FALSE}
    .attr("width", (row) => row.value);
```

* set `<rect>` width to "value" column

Nice! Now let's try with some bigger values:

```{r echo=FALSE}
dummy_data <- data.frame(
  category = c("A", "B", "C"),
  value = c(3000, 1000, 500)
)
knitr::kable(dummy_data)
```

<div id="plot-parent-bigger"></div>

```{js echo=FALSE}
dummy_data = [
  { category: "A", value: 3000 },
  { category: "B", value: 1000 },
  { category: "C", value: 500 }
];

my_svg = initSvg("#plot-parent-bigger");
addDummyRect(my_svg);
```

Ok. Now we realy need scale. `d3.scaleLinear` takes two arguments - domain and range:

* domain - minimum and maximum value in our data
* range - minimum and maximum value that we want map to (width of our svg in this case)

```{js eval=FALSE}
const svgWidth = 500;
const valueMax = d3.max(dummy_data, (row) => row.value);

const scaleX = d3.scaleLinear(
  [0, valueMax],
  [0, svgWidth]
)
```

*"Ok, but what's that???"*

```{js eval=FALSE}
d3.max(dummy_data, (row) => row.value);
```

*"Why can't you just do something like:"*

```{js eval=FALSE}
d3.max(dummy_data.value);
```

In JS "data.frame" is an array of objects:

```{js eval=FALSE}
[
  { category: "A", value: 3000 },
  { category: "B", value: 1000 },
  { category: "C", value: 500 }
];
```

We can compare it to a list of named lists in R:

```{r, eval=FALSE}
list(
  list(category = "A", value = 3000),
  list(category = "B", value = 1000),
  list(category = "C", value = 500)
)
```

So to select some "column" in JS we need to map through array objects:

```{js, eval=FALSE}
dummy_data.map((row) => row.value);
```

And `d3.min` function takes accessor function as second parameter.

Back to our scale. `d3.scaleLinear` returns a function that maps our data values (domain) to pixels on svg (range).

Equivalent function in R would look like this:

```{r}
scale_linear <- function(domain, range) {
  function(x) {
    (x - min(domain)) / (max(domain) - min(domain)) * (max(range) - min(range)) + min(range)
  }
}
```

Now let's try to plot out dataset with bigger values using scale:

```{js eval=FALSE}
  .attr("width", (row) => scaleX(row.value));
```

<div id="plot-parent-bigger-scaled"></div>

```{js echo=FALSE}
let svgWidth = 500;
let valueMax = d3.max(dummy_data, (row) => row.value);

let scaleX = d3.scaleLinear(
  [0, valueMax],
  [0, svgWidth]
)

my_svg = initSvg("#plot-parent-bigger-scaled");
addDummyRect(my_svg, scaleX);
```

We must add axis to actualy see the efect.






<br/>
<br/>
<br/>
<br/>
<br/>
<br/>
<br/>




Probably we will also need some margins for chart axes:

```{js eval=FALSE}
const width = 500;
const height = 300;
const margin = {
  top: 10,
  right: 10,
  bottom: 30,
  left: 50
}
const panelWidth = width - margin.left - margin.right;
const panelHeight = height - margin.top - margin.bottom;

const svg = d3.select("#plot-parent")
  .append("svg")
  .attr("width", 500)
  .attr("height", 300)


```



