---
title: "D3 tutorial for R users - barplot"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(shiny)
library(purrr)
```

<script src="www/d3.min.js" />

```{js echo=FALSE}
const initSvg = (id) => {
  const mySvg = d3
    .select(id)
    .append("svg")
    .attr("width", 500)
    .attr("height", 200);
  return mySvg;
}
const addDummyRect = (svg, scaleX = (x) => x, height = 200) => {
  svg.selectAll(".my-bar-class")
    .data(dummyData)
    .join("rect")
    .attr("x", 0)
    .attr("height", height / 3 * 0.8)
      .attr("y", (_, i) => height / 3 * i)
      .attr("width", (row) => scaleX(row.value));
}
```

> **"I can't learn D3 because I don't know JavaScript."** - `FALSE`

Some knowledge of JavaScript is needed, but really very basic. Learning new language is just remembering some conventions and code syntax. Most important is understanding what code is doing. Remebering the syntax will come with time.

# svg

To make chart `<svg>` element is needed. By default it has size of 300*x*150 but we can set it with `width` and `height` attributes.

With D3.js we can select our chart parent element and append `svg` to it.

```{js eval=FALSE}
const mySvg = d3
  .select("#plot-parent")
  .append("svg")
  .attr("width", 500)
  .attr("height", 200);
```

(`select`method accepts any css selector)

We can achieve the same result with Shiny:

```{r eval=FALSE}
tags$svg(width = 500, height = 200)
```

Then in D3 we would olny need to store created `<svg>` in a variable:

```{js eval=FALSE}
const mySvg = d3.select("svg");
```

Since we're Shiny users we will proceed with second option.

# Bars

Now we need to add some margin, scales, axes... Ok let's put some bars first to not get too bored at the beginning.

We will use following dummy data:

```{r echo=FALSE}
dummy_data <- data.frame(
  category = c("A", "B", "C"),
  value = c(300, 100, 50)
)
knitr::kable(dummy_data)
```

We can add bars with the following code:

```{js eval=FALSE}
mySvg.selectAll(".my-bar-class")
  .data(dummyData)
  .join("rect")
    .attr("x", 0)
    .attr("height", 200 / 3 * 0.8)
    .attr("y", (_, i) => 200 / 3 * i)
    .attr("width", (row) => row.value);
```

<div id="plot-parent"></div>

```{js echo=FALSE}
let dummyData = [
  { category: "A", value: 300 },
  { category: "B", value: 100 },
  { category: "C", value: 50 }
];

let mySvg = initSvg("#plot-parent");
addDummyRect(mySvg);
```

*"Ok... we have 3 ugly rectangles and some spaghetti code that is not easy at all."*
Yee, looks a bit strange, but let's try to understand what this code do. To make it simpler I will do the same using R:

```{r}
tags$svg(
  width = 500, height = 200,
  map(1:nrow(dummy_data), function(i) {
    tags$rect(
      x = 0,
      height = 200 / 3 * 0.8,
      y = 200 / 3 * (i - 1),
      width = dummy_data[i, ]$value
    )
  })
)
```

The main thing this code does is iterating through data and map it to `<rect>` elements. Let's also discuss D3 code in details:

```{js eval=FALSE}
mySvg.selectAll(".my-bar-class")
```

* since there are no elements with class ".my-bar-class" in our svg, D3 creates placeholders (for data and elements)

```{js eval=FALSE}
  .data(dummyData)
```

* join `dummyData` to placeholders

```{js eval=FALSE}
  .join("rect")
```

* append 3 `<rect>` elements (1 for each data row)

```{js eval=FALSE}
    .attr("x", 0)
```

* our bars starts on x = 0

```{js eval=FALSE}
    .attr("height", 200 / 3 * 0.8)
```

* height is calculated by dividing plot height by number of bars and multiply by 0.8 to give some padding

```{js eval=FALSE}
    .attr("y", (row, i) => 200 / 3 * i)
```

* in `attr` method we can iterate over data (that we joined earlier with `data` method) using function - first argument is data row, second is row number

  I use *arrow function* for brevity but we can use regular function as well - `function(row, i) { ... }`

```{js eval=FALSE}
    .attr("width", (row) => row.value);
```

* set `<rect>` width to "value" column

# Scale

Now let's try with smaller values:

```{r echo=FALSE}
dummy_data <- data.frame(
  category = c("A", "B", "C"),
  value = c(30, 10, 5)
)
knitr::kable(dummy_data)
```

<div id="plot-parent-bigger"></div>

```{js echo=FALSE}
dummyData = [
  { category: "A", value: 30 },
  { category: "B", value: 10 },
  { category: "C", value: 5 }
];

mySvg = initSvg("#plot-parent-bigger");
addDummyRect(mySvg);
```

Ok. Now we realy need scale. `d3.scaleLinear` takes two arguments - domain and range:

* domain - minimum and maximum value in our data
* range - minimum and maximum value that we want map to (width of our svg in this case)

```{js eval=FALSE}
const svgWidth = 500;
const valueMax = d3.max(dummyData, (row) => row.value);

const scaleX = d3.scaleLinear(
  [0, valueMax],
  [0, svgWidth]
);
```

*"Ok, but what's that???"*

```{js eval=FALSE}
d3.max(dummyData, (row) => row.value);
```

*"Why can't you just do something like:"*

```{js eval=FALSE}
d3.max(dummyData.row);
```

In JS "data.frame" is an array of objects:

```{js eval=FALSE}
[
  { category: "A", value: 30 },
  { category: "B", value: 10 },
  { category: "C", value: 5 }
];
```

We can compare it to a list of named lists in R:

```{r, eval=FALSE}
list(
  list(category = "A", value = 30),
  list(category = "B", value = 10),
  list(category = "C", value = 5)
)
```

So, to select some "column" in JS we need to map through array objects and pull given value from each:

```{js, eval=FALSE}
dummyData((row) => row.value);
```

`d3.max` function takes accessor function as second parameter.

Back to our scale. `d3.scaleLinear` returns a function that maps our data values (domain) to pixels on svg (range).

Equivalent function in R would look like this:

```{r}
scale_linear <- function(domain, range) {
  function(x) {
    (x - min(domain)) / (max(domain) - min(domain)) * (max(range) - min(range)) + min(range)
  }
}
```

If we plot our new data using scale the biggest bar will fill the whole `<svg>`:

```{js eval=FALSE}
  .attr("width", (row) => scaleX(row.value));
```

<div id="plot-parent-bigger-scaled"></div>

```{js echo=FALSE}
let svgWidth = 500;
let valueMax = d3.max(dummyData, (row) => row.value);

let scaleX = d3.scaleLinear(
  [0, valueMax],
  [0, svgWidth]
);

mySvg = initSvg("#plot-parent-bigger-scaled");
addDummyRect(mySvg, scaleX);
```

# Axis

To see if we realy have correct values we must add axis. This time D3 offers very convinient function:

In Shiny we have to add `<g>` tag to append our axis to:

```{r eval=FALSE}
tags$svg(
  width = 500, height = 200,
  tags$g(class = "axis-x")
)
```

In JS we have to create axis by passing our scale to `axisBottom` function, select our `<g>` tag with class `"axis-x"` and append axis to it.

```{js eval=FALSE}
const axisX = d3.axisBottom(scaleX);
mySvg.select(".axis-x").call(axisX);
```

`<g>` tag is similar to `<div>` with one difference - it can't have any visual style. It's used only to group other elements.
`call` method is used to call `axisX` as function - `axisX()` and append result to `<g>` tag.

<div id="plot-axis"></div>

```{js echo=FALSE}
mySvg = initSvg("#plot-axis");
addDummyRect(mySvg, scaleX);
let axisX = d3.axisBottom(scaleX);
mySvg.append("g").call(axisX);
```

```{css echo=FALSE}
#plot-axis text {
  fill:white
}
#plot-axis line {
  stroke:white
}
```

*"Didn't I said axis**Bottom**??? Why it's on top?"*

Axis bottom means that labels and ticks are on the bottom of the axis. D3 is not aweare what's your chart height, so it put's it at (0, 0) coordinates. It's our duty to put it in right place. We can do it using `transform` attribute:

```{js eval=FALSE}
const svgHeight = 200;

mySvg.select(".axis-x")
  .attr("transform", `translate(0, ${svgHeight})`)
  .call(axisX);
```

Here - `` `translate(0, ${svgHeight})` `` - I used string interpolation, similar to R `glue`. In JS this functionality is built in. You have to use backticks - `` ` ``, and inside `${}` you can put variable or code that will evaluate.

<div id="plot-axis-transform"></div>

```{js echo=FALSE}
const svgHeight = 200;
mySvg = initSvg("#plot-axis-transform");
addDummyRect(mySvg, scaleX);
mySvg.append("g")
  .attr("transform", `translate(0, ${svgHeight})`)
  .call(axisX);
```

## Margin

Our axis is this thin line at the bottom. We can't see ticks and labels because they are outside of our svg. To make them visible we need to add margins to out chart. To do this we can use `<g>` tag and `transform` attribute.

```{r eval=FALSE}
tags$svg(
  width = 500, height = 200,
  tags$g(
    class = "plot",
    tags$g(class = "axis-x")
  )
)
```

```{js eval=FALSE}
const svgWidth = 500;
const svgHeight = 200;
const margin = {
  top: 10,
  right: 10,
  bottom: 30,
  left:30
};
const plotWidth = svgWidth - margin.left - margin.right;
const plotHeight = svgHeight - margin.top - margin.bottom;

const mySvg = d3
  .select("svg")
  .attr("width", svgWidth)
  .attr("height", svgHeight);

const plot = mySvg
  .select(".plot")
  .attr("transform", `translate(${margin.left}, ${margin.top})`);
```

To better visualize what we achieved I've added black border to `<svg>` element and rectangle with blue border that represents `plot` element dimensions and position. Now all elements appended to `plot` will have origin in the point marked with blue dot.

<svg width="500" height="200">
  <rect width="500" height="200" style="fill: white; stroke: black"></rect>
  <g transform="translate(30, 10)">
    <rect width="460" height="160" style="fill: white; stroke: steelblue"></rect>
    <circle r="4" fill="steelblue" />
    <text x="5" y="15">(0, 0)</text>
  </g>
</svg>

With respect to the blue point x-axis has to be `transform`ed in y-dimension with a value of `plotHeight`.

We also have to append `<rect>` elements to `plot`.

```{js eval=FALSE}
plot.select(".axis-x")
  .attr("transform", `translate(0, ${plotHeight})`)
  .call(axisX);

plot.selectAll(".my-bar-class")
  .data(dummyData)
  .join("rect")
    .attr("x", 0)
    .attr("height", plotHeight / 3 * 0.8)
    .attr("y", (_, i) => plotHeight / 3 * i)
    .attr("width", (row) => scaleX(row.value));
```

<svg id="plot-axis-margin">
  <g class="plot">
    <g class="axis-x"></g>
  </g>
</svg>

```{js echo=FALSE}
const margin = {
  top: 10,
  right: 10,
  bottom: 30,
  left:30
};
const plotWidth = svgWidth - margin.left - margin.right;
const plotHeight = svgHeight - margin.top - margin.bottom;

mySvg = d3
  .select("#plot-axis-margin")
  .attr("width", svgWidth)
  .attr("height", svgHeight);

let plot = mySvg
  .select(".plot")
  .attr("transform", `translate(${margin.left}, ${margin.top})`);

scaleX.range([0, plotWidth]);

addDummyRect(plot, scaleX, plotHeight);
plot.select(".axis-x")
  .attr("transform", `translate(0, ${plotHeight})`)
  .call(axisX);
```
Axis ticks and labels are now visible. We can also see that bars width is scaled properly.

The last piece we're missing is y-axis. Let's add it now. For categorical variables most often `d3.scaleBand` is used. As a domain it accepts array of unique values.

```{js eval=FALSE}
const domainY = [...new Set(dummyData.map((row) => row.category))];
const scaleY = d3
  .scaleBand(domainY, [0, plotHeight])
  .padding(0.2);
const axisY = d3.axisLeft(scaleY);
plot.select(".axis-x").call(axisY);

plot.selectAll(".my-bar-class")
  .data(dummyData)
  .join("rect")
    .attr("x", 0)
    .attr("height", scaleY.bandwidth())
    .attr("y", (row) => scaleY(row.category))
    .attr("width", (row) => scaleX(row.value));
```

* `[...new Set(dummyArray)]` is equivalent to `unique(dummy_vector)`
* with `padding` method we can get space around bars (value returned by `scaleY.bandwidth()` depends on value passed to `padding`)
* y-axis should start at point (0, 0) of our `plot` so `transform` is not needed here
* we can use `scaleY` to get `y` position of bars

<svg id="plot-complete">
  <g class="plot">
    <g class="axis-x"></g>
  </g>
</svg>

```{js echo=FALSE}
mySvg = d3
  .select("#plot-complete")
  .attr("width", svgWidth)
  .attr("height", svgHeight);

plot = mySvg
  .select(".plot")
  .attr("transform", `translate(${margin.left}, ${margin.top})`);

plot.select(".axis-x")
  .attr("transform", `translate(0, ${plotHeight})`)
  .call(axisX);

const domainY = [...new Set(dummyData.map((row) => row.category))];
const scaleY = d3.scaleBand(domainY, [0, plotHeight]).padding(0.2);
const axisY = d3.axisLeft(scaleY);
plot.append("g").call(axisY);

plot.selectAll(".my-bar-class")
  .data(dummyData)
  .join("rect")
    .attr("x", 0)
    .attr("height", scaleY.bandwidth())
    .attr("y", (row) => scaleY(row.category))
    .attr("width", (row) => scaleX(row.value));
```

# Pimp my chart

Default D3 chars doesn't look good. It's possible to change styles with D3 but personally I prefer to use CSS.

## Axis

D3 axis consists of 3 elements:

* <span style="color: #50CB93">domain line</span>
* <span style="color: #ED8E7C">ticks</span>
* <span style="color: #A03C78">labels</span>

We can change them with following CSS selectors:

```{css eval=FALSE}
.domain {
  stroke: #50CB93;
}
.tick > line {
  stroke: #ED8E7C;
}
.tick > text {
  color: #A03C78;
}
```

<svg id="axis-style" width="700" height="40"></svg>

```{js echo=FALSE}
d3.select("#axis-style").append("g").attr("transform", "translate(15, 10) scale(1.3)").call(axisX);
```

```{css echo=FALSE}
#axis-style .domain {
  stroke: #50CB93;
}
#axis-style .tick > line {
  stroke: #ED8E7C;
}
#axis-style .tick > text {
  fill: #A03C78;
}
```

## Grid lines

To get grid lines with minimal effort I often use the trick to set axis ticks size to negative plot width/height.

```{js eval=FALSE}
const axisX = d3.axisBottom(scaleX).tickSize(-plotHeight);
const axisY = d3.axisLeft(scaleY).tick(-plotWidth);

plot.select(".axis-x")
  .attr("transform", `translate(0, ${plotHeight})`)
  .call(axisX)
plot.select(".axis-y")
  .call(axisY);
```

<svg id="plot-grid">
  <g class="plot">
    <g class="axis-x"></g>
    <g class="axis-y"></g>
  </g>
</svg>

```{js echo=FALSE}
mySvg = d3
  .select("#plot-grid")
  .attr("width", svgWidth)
  .attr("height", svgHeight);

plot = mySvg
  .select(".plot")
  .attr("transform", `translate(${margin.left}, ${margin.top})`);

plot.select(".axis-x")
  .attr("transform", `translate(0, ${plotHeight})`)
  .call(axisX.tickSize(-plotHeight));

plot.select(".axis-y")
  .call(axisY.tickSize(-plotWidth));

plot.selectAll(".my-bar-class")
  .data(dummyData)
  .join("rect")
    .attr("x", 0)
    .attr("height", scaleY.bandwidth())
    .attr("y", (row) => scaleY(row.category))
    .attr("width", (row) => scaleX(row.value));
```

## CSS example

* remove both domain lines and y-axis ticks

```{css eval=FALSE}
.domain, .axis-y > .tick > line {
  stroke: transparent;
}
```

* make x-axis grid lines lighter

```{css eval=FALSE}
.axis-x > .tick > line {
  opacity: 0.2;
}
```

* change axes labels font and size

```{css eval=FALSE}
.tick > text {
  font-family: Verdana;
  font-size: 12px;
}
```

* move y-axis labels a bit further from the bars

```{css eval=FALSE}
.axis-y > .tick > text {
  transform: translateX(-10px);
}
```

* change the color of bars and setting some opacity

```{css eval=FALSE}
rect {
  fill: #125D98;
  opacity: 0.6;
}
```

<div id="plot-style" class="plot-style"></div>

```{js echo=FALSE}
$("#plot-style").append($("#plot-grid").clone());
```

# Transitions

It was a lot of effort to make a simple barplot. But now, adding smooth, good looking transitions to our plot is super easy!

```{js eval=FALSE}
plot.select(".axis-x")
  .attr("transform", `translate(0, ${plotHeight})`)
  .transition()
  .call(axisX);

plot.select(".axis-y")
  .transition()
  .call(axisY);

plot.selectAll(".my-bar-class")
  .data(dummyData)
  .join(
    (enter) => {
      enter
        .append("rect")
        .attr("class", "my-bar-class")
        .attr("x", 0)
        .attr("height", scaleY.bandwidth())
        .attr("y", (row) => scaleY(row.category))
        .attr("width", (row) => scaleX(row.value));
    },
    (update) => {
      update
        .transition()
        .attr("width", (row) => scaleX(row.value));
    }
  );
```

On axes `<g>` tags we have to call `transition` method before `call`ing axis.
In case of `<rect>` elements also only adding `transition` method before changeing attributes is necessary, but we also have to handle differently two situations:

* `enter` - when `<rect>` elements are created
* `update` - when `<rect>` elements are updated
* `exit` - there is also 3rd case, when element is removed from the plot, but it will now be needed in our case

`join` method is overloaded - can accept:

* element name (single string eg - "rect") - then it just `append` element to placeholders
* 1-3 functions (`(enter) => {...}`, `(update) => {...}`, `(remove) => {...}`) - then it applies one of 3 functions to selection elements based on the element status

We also have to add to the `<rect>` elements the same class that we use in `selectAll` (`"my-bar-class"`). Otherwise D3 will add new elements on each update.

That's it! Thanks to magical `transition` method we can repeat after Dr Frankensten *"It's alive!"*

<button onclick=updateData()>It's alive!</button>

<div class="plot-style">
  <svg id="plot-interactive">
    <g class="plot">
      <g class="axis-x"></g>
      <g class="axis-y"></g>
    </g>
  </svg>
</div>

```{js echo=FALSE}
let randomData = dummyData;

let updateData = () => {
  randomData.forEach((row) => {
    row.value = Math.random() * 100;
  });
  updatePlot(randomData);
}

mySvg = d3
  .select("#plot-interactive")
  .attr("width", svgWidth)
  .attr("height", svgHeight)

plot = mySvg
  .select(".plot")
  .attr("transform", `translate(${margin.left}, ${margin.top})`);

plot.select(".axis-y").call(axisY);

let updatePlot = (data) => {
  let valueMax = d3.max(data, (row) => row.value);
  scaleX.domain([0, valueMax]);
  plot.select(".axis-x")
    .attr("transform", `translate(0, ${plotHeight})`)
    .transition()
    .call(axisX)

  plot.selectAll(".my-bar-class")
    .data(data)
    .join(
      (enter) => {
        enter
          .append("rect")
          .attr("class", "my-bar-class")
          .attr("x", 0)
          .attr("height", scaleY.bandwidth())
          .attr("y", (row) => scaleY(row.category))
          .attr("width", (row) => scaleX(row.value));
      },
      (update) => {
        update
          .transition()
          .attr("width", (row) => scaleX(row.value));
      }
    );
};

updatePlot(randomData);
```

```{css echo=FALSE}
.plot-style .domain, .plot-style .axis-y > .tick > line {
  stroke: transparent;
}
.plot-style .axis-x > .tick > line {
  opacity: 0.2;
}
.plot-style .tick > text {
  font-family: Verdana;
  font-size: 12px;
}
.plot-style .axis-y > .tick > text {
  transform: translateX(-10px);
}
.plot-style rect {
  fill: #125D98;
  opacity: 0.6;
}
```

# Wrap it up

Final working example of Shiny app with D3 barplot is:

* `app.R`

```{r eval=FALSE}
library(shiny)
library(purrr)

ui <- fluidPage(
  tags$head(
    tags$link(rel = "stylesheet", type = "text/css", href = "style.css"),
    tags$script(src = "index.js"),
    tags$script(src = "d3.min.js")
  ),
  div(
    class = "card",
    actionButton(inputId = "update", label = "It's alive!"),
    tags$svg(
      tags$g(
        class = "plot",
        tags$g(class = "axis-x"),
        tags$g(class = "axis-y")
      )
    )
  )
)

server <- function(input, output, session) {
  generate_data <- function() {
    data.frame(
      category = c("A", "B", "C"),
      value = runif(3) * 100
    )
  }
  render_barplot <- function() {
    session$sendCustomMessage("render_barplot", transpose(generate_data()))
  }
  render_barplot()
  observeEvent(input$update, {
    render_barplot()
  })
}

shinyApp(ui, server)
```

* `www/index.js`

```{js eval=FALSE}
$(function () {
  Shiny.addCustomMessageHandler("render_barplot", function (dummyData) {
    const svgWidth = 500;
    const svgHeight = 200;
    const margin = {
      top: 30,
      right: 10,
      bottom: 30,
      left: 30,
    };
    const plotWidth = svgWidth - margin.left - margin.right;
    const plotHeight = svgHeight - margin.top - margin.bottom;

    const domainX = [0, d3.max(dummyData, (row) => row.value)];
    const domainY = [...new Set(dummyData.map((row) => row.category))];

    const scaleX = d3.scaleLinear(domainX, [0, plotWidth]);
    const scaleY = d3.scaleBand(domainY, [0, plotHeight]).padding(0.2);

    const axisX = d3.axisBottom(scaleX).tickSize(-plotHeight);
    const axisY = d3.axisLeft(scaleY);

    mySvg = d3.select("svg").attr("width", svgWidth).attr("height", svgHeight);

    plot = mySvg
      .select(".plot")
      .attr("transform", `translate(${margin.left}, ${margin.top})`);

    plot
      .select(".axis-x")
      .attr("transform", `translate(0, ${plotHeight})`)
      .transition()
      .call(axisX);

    plot.select(".axis-y").call(axisY);

    plot
      .selectAll(".bar")
      .data(dummyData)
      .join(
        (enter) => {
          enter
            .append("rect")
            .attr("class", "bar")
            .attr("x", 0)
            .attr("height", scaleY.bandwidth())
            .attr("y", (row) => scaleY(row.category))
            .attr("width", (row) => scaleX(row.value));
        },
        (update) => {
          update.transition().attr("width", (row) => scaleX(row.value));
        }
      );
  });
});
```

* `www/style.css`

```{css eval=FALSE}
.domain, .axis-y > .tick > line {
  stroke: transparent;
}
.axis-x > .tick > line {
  opacity: 0.2;
}
.tick > text {
  font-family: Verdana;
  font-size: 12px;
}
.axis-y > .tick > text {
  transform: translateX(-10px);
}
rect {
  fill: #125D98;
  opacity: 0.6;
}

/* card style */
.card {
  display: flex;
  flex-direction: column;
  justify-content: center;
  align-items: center;
  border: 0.1px solid lightgray;
  box-shadow: 5px 5px 5px lightgray;
  width: min-content;
  padding: 40px;
}
.container-fluid {
  display: flex;
  justify-content: center;
  margin-top: 50px;
}
```

